---
title: "Project-HRH"
author: "Alesio, Audrey, and Ian"
date: "2/8/2024"
output: html_document
---

```{r}
library(tidyverse)
library(ggplot2)
```

```{r}
data <- read.csv("2023HRH.csv", check.names = FALSE, fileEncoding = "Latin1", na.strings="")
colnames(data) <- make.names(colnames(data), unique = TRUE)
```

```{r}
#Creates a list of all variables being removed from the dataset (contain only NA values)
NA_vars <- data[,colSums(is.na(data))==nrow(data)]
NA_var_names <- colnames(NA_vars)
NA_var_names
```

```{r}
#Filter the dataset to exclude any columns with only NA values
data <- data[,colSums(is.na(data))<nrow(data)]
```

```{r Variables 1-15 (Ian)}
data %>% rename(info.consent = Info.Sharing.Level) %>% factor(info.consent, levels = c("I agree to allow my information to be shared inside the CRHMIS for coordination of care purposes.<ul><li>a) I agree to allow my information to be shared with external sources for research purposes.</li></ul>"))
```

```{r Variables 16-30 (Alesio)}

```


```{r Variables 31-45 (Audrey)}
#each pair of newname = oldname is separated by a comma
data <- data %>% rename(TO.AfghanistanFreedom = Theatre.of.Operations..Afghanistan..Operation.Enduring.Freedom.,
                        TO.IraqiFreedom = Theatre.of.Operations..Iraq..Operation.Iraqi.Freedom.,
                        TO.IraqNewDawn = Theatre.of.Operations..Iraq..Operation.New.Dawn.,
                        TO.KoreanWar = Theatre.of.Operations..Korean.War,
                        TO.Other = Theatre.of.Operations..Other.Peace.keeping.Operations.or.Military.Interventions,
                        TO.PersianGulf = Theatre.of.Operations..Persian.Gulf.War..Operation.Desert.Storm.,
                        TO.Vietnam = Theatre.of.Operations..Vietnam.War,
                        TO.WW2 = Theatre.of.Operations..World.War.II)

data <- data %>% rename(months.homeless.3y = Total.number.of.months.homeless.on.the.street..in.ES..or.SH.in.the.past.three.years..Admission.,
                       res.type = Type.of.Residence..Admission.,
                       vet.discharge.status = Veteran.Discharge.Status,
                       vet.status = Veteran.Status,
                       apprentice = Vocational.Training.Apprentice,
                       mil.year.start = Year.Entered.Military.Service,
                       mil.year.end = Year.Separated.from.Military.Service,
                       last.zipcode = Zip.Code.of.Last.Permanent.Address)
```


```{r output new csv}
#unsure if possible to export directly to the github repo. here you need to change
#the code to the desired location on your computer 

#write.csv(data, file='/Users/user/Desktop/new_file.csv', row.names=FALSE)
```


```{r Missing data}
missingcount <- colSums(is.na(data)) 
missingprop <- missingcount/nrow(data)
missingdata <- cbind(missingcount, missingprop)
colnames(missingdata) <- c("NA_count", "NA_proportion")
```

```{r Missing Data Example}
data %>% count(If.the.person.is.60.years.of.age.or.older..score.1)
data %>% count(Current.Age >= 60)

data %>% filter(Current.Age >= 60) %>% filter(is.na(If.the.person.is.60.years.of.age.or.older..score.1))

data %>% filter(Current.Age >= 60) %>% filter(If.the.person.is.60.years.of.age.or.older..score.1 == 0) %>% count(Program, sort = TRUE)

data %>% filter(Current.Age >= 60) %>% filter(is.na(If.the.person.is.60.years.of.age.or.older..score.1) | If.the.person.is.60.years.of.age.or.older..score.1 == 0) %>% count(Program, sort = TRUE)
```


```{r plot function}
#inputs: df: dataframe containing predictor and outcome
#     colindexpred: index of predictor variable (categorical)
#     colindexout: index of outcome variable (numeric)
#     includena: default true, excludes na values if FALSE
#note- relies on missing data proportions being saved in missingprop, 
#     where the cols are in same order as in df

EDAplot <- function(df, colindexpred, colindexout, includena=TRUE){
  mytitle <- paste(colnames(df)[colindexout], " vs. ", colnames(df)[colindexpred])
  propvalue <- paste("Proportion of observations/total rows: ",
                     round(1-missingprop[colindexpred], digits=3))
  
  df <- df[,c(colindexpred,colindexout)]
  if (includena == FALSE){df <- na.omit(df)}
  
  myplot <- ggplot(data = df, aes(x = as.factor(df[,1]), y=df[,2])) + 
    geom_boxplot() +       
    theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))+
    ggtitle(mytitle, subtitle = propvalue)
  print(myplot)
}
```

```{r demographic plots 51-104}
ggplot(data, aes(x=Gender..HMIS.)) + geom_bar() + 
  ggtitle("Gender of Clients in System") +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))

ggplot(data, aes(x= reorder(HMIS.Project.Type, HMIS.Project.Type, FUN=length))) +
  geom_bar() +  ggtitle("Project Type") + xlab("HMIS Project Type") +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))

ggplot(data, aes(x=HMIS.Household.Type..at.admission.)) + 
  geom_bar() + ggtitle("Household Type at Admission", subtitle = round(1-missingdata["HMIS.Household.Type..at.admission.",2],digits=3)) +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))

ggplot(data, aes(x=Length.of.Program.Stay, fill = Program.Type)) + geom_histogram(bins=30) + ggtitle("Length of Program Stay by Program Type")

ggplot(data, aes(x=Mental.Health.Disorder..Admission.)) + geom_bar() + 
  ggtitle("Mental Health Disorder", 
          subtitle = round(1-missingdata["Mental.Health.Disorder..Admission.",2],digits=3)) +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))

ggplot(data, aes(x=reorder(Length.of.Stay.in.Prior.Living.Situation..Admission.,Length.of.Stay.in.Prior.Living.Situation..Admission.,FUN=length))) + 
  geom_bar() + 
  ggtitle("Length of Stay in Prior Living Situation") +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))

```

```{r demographic plots 1-50}
ggplot(data.frame(data), aes(x=Did.you.stay.less.than.7.nights...Admission.)) + geom_bar()

ggplot(data.frame(data), aes(x=Did.you.stay.less.than.90.days...Admission.)) + geom_bar()

box_plot <- ggplot(data, aes(x=Did.you.stay.less.than.7.nights...Admission., y = Length.of.Program.Stay)) 
box_plot + geom_boxplot()

ggplot(data, aes(x=Age.at.Admission)) + geom_bar() + ggtitle("Age of Clients at Admission") +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))

barplot(data$Length.of.Program.Stay, names=data$Admission.Date...Project.Start.Date,
        xlab = "Project Start Time",
        ylab = "Length Of Stay")


ggplot(data, aes(x=Chronically.Homeless..Admission.)) + geom_bar() + ggtitle("Chronically Homeless Clients in System") +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))

box_plot <- ggplot(data, aes(x=Ethnicity, y = Length.of.Program.Stay))
box_plot + geom_boxplot()

data$Admission.Date...Project.Start.Date <- as.Date(data$Admission.Date...Project.Start.Date)
ggplot(data, aes(x=Admission.Date...Project.Start.Date, 
       y = Length.of.Program.Stay)) + geom_point(aes(color = Admission.Date...Project.Start.Date)) +
scale_color_continuous("Project Start Dates") +
xlab("Project Start Time") +
ylab("Length Of Stay") +
  theme(axis.line = element_line(color = "black", size = 0.5))

```

```{r Updated Demographic Plots}

ggplot(data, aes(x=Length.of.Program.Stay, fill = Program.Type)) + geom_histogram(bins=30) + ggtitle("Length of Program Stay by Program Type")

ggplot(data, aes(x=Age.at.Admission, fill = Program.Type)) + geom_histogram(bins=30) + ggtitle("Age of Clients by Program Type")

ggplot(data, aes(x=Age.at.Admission, y = Program.Type)) + geom_point() + ggtitle("Age of Clients by Program Type") 

ggplot(data, aes(x= reorder(HMIS.Project.Type, HMIS.Project.Type, FUN=length))) +
  geom_bar() +  ggtitle("Project Type") + xlab("HMIS Project Type") +
  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))

ggplot(data, aes(Did.you.stay.less.than.7.nights...Admission., ..count..)) + geom_bar(aes(fill = Program.Type), position = "dodge")

ggplot(data, aes(Did.you.stay.less.than.90.days...Admission., ..count..)) + geom_bar(aes(fill = Program.Type), position = "dodge")

ggplot(data, aes(Gender..HMIS., ..count..)) + geom_bar(aes(fill = Program.Type), position = "dodge")

ggplot(data, aes(x=Age.at.Admission, fill = Program.Type)) + geom_bar() + ggtitle("Age of Clients by Program Type") + facet_wrap(~Program.Type)

ggplot(data, aes(x=Length.of.Program.Stay, fill = Program.Type)) + geom_histogram() + ggtitle("Length of Program Stay by Program Type") + facet_wrap(~Program.Type)
```
```{r Density Plots}
ggplot(data, aes(x=Length.of.Program.Stay, fill = Program.Type)) + geom_density() + ggtitle("Length of Program Stay by Program Type") + facet_wrap(~Program.Type) + theme(axis.text.x = element_text(angle = 45,vjust = 0.25, hjust = 0.25))

ggplot(data, aes(x=Length.of.Program.Stay, fill = Program.Type)) + geom_density()+ ggtitle("Length of Program Stay by Program Type")

ggplot(data, aes(x=Age.at.Admission, fill = Program.Type)) + geom_density() + ggtitle("Age of Clients by Program Type") + facet_wrap(~Program.Type)

```

```

```{r recode race and plot}
data$Race..HMIS. <- 
  fct_collapse(data$Race..HMIS., 
               Black = "Black, African American, or African", 
               N_A = c("Client refused", "Data not collected", ""),
               AmericanIndian_AlaskanNaitive_Indegenous = "American Indian, Alaska Native, or Indigenous", 
               White = "White", 
               Hawaiian_PacificIslander = "Native Hawaiian or Pacific Islander", 
               Asian = "Asian or Asian American", 
               Hispanic_Latinx = "Hispanic/Latina/e/o",
               White_HispLatinx = c("White;Hispanic/Latina/e/o", "Hispanic/Latina/e/o;White"), 
               Black_HispLatinx = "Black, African American, or African;Hispanic/Latina/e/o", 
               other_level="2+ Races")

data %>% ggplot(aes(x = Race..HMIS., y = Length.of.Program.Stay)) + geom_boxplot()+  theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))

data %>% ggplot(aes(x = Race..HMIS.)) + geom_bar()+theme(axis.text.x = element_text(angle = 45,vjust = 1, hjust = 1))
```

```{r Program-specific table}
programinfo <- data %>%  group_by(HMIS.Project.Type, Program.Type, Program) %>% summarize(n = n(), MedianLengthofStay = median(Length.of.Program.Stay))
```

```{r demographic plots 100+}
#are there patterns between incompletely collected data and length of stay?

ggplot(data = data, aes(x = On.the.night.before.did.you.stay.on.the.streets..ES.or.SH...Admission., y = Length.of.Program.Stay)) + geom_violin()

#data %>% count(Other..Admission.)
#Healthcare Variable

ggplot(data = data, aes(x = Person.Type..at.admission., y = Length.of.Program.Stay)) + geom_boxplot()

data %>% ggplot(aes(x = Post.Secondary.Education, y = Length.of.Program.Stay)) + geom_boxplot()

#data %>% ggplot(aes(x = Preferred.Language, y = Length.of.Program.Stay)) + geom_boxplot()

data %>% ggplot(aes(x = Previous.Admission., y = Length.of.Program.Stay)) + geom_boxplot()

data %>% ggplot(aes(x = Program.Type, y = Length.of.Program.Stay)) + geom_boxplot()

#for program type, plot average length of stay across each program type

data %>% group_by(Program) %>% summarise_at(vars(Length.of.Program.Stay), list(mean = mean)) %>% ggplot(aes(x = Program, y = mean)) + geom_point()

data %>% ggplot(aes(x = Race..HMIS., y = Length.of.Program.Stay)) + geom_boxplot()

data %>% count(Services.Sought)

data %>% count(Sexual.Orientation)

data %>% ggplot(aes(x = Substance.Use.Disorder..Admission., y = Length.of.Program.Stay)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90))

data %>% count(Total.number.of.months.homeless.on.the.street..in.ES..or.SH.in.the.past.three.years..Admission.)
      
data %>% ggplot(aes(x = Total.number.of.months.homeless.on.the.street..in.ES..or.SH.in.the.past.three.years..Admission., y = Length.of.Program.Stay)) + geom_boxplot() + theme(axis.text.x = element_text(angle = 90))

mutate(data, Race..HMIS.Cat = recode(Race..HMIS., Black = "Black, African American, or African", N_A = c("Client refused", "Data not collected", ""),AmericanIndian_AlaskanNaitive_Indegenous = "American Indian, Alaska Native, or Indigenous", White = "White", Hawaiian_PacificIslander = "Native Hawaiian or Pacific Islander", Asian = "Asian or Asian American", Hispanic/Latinx = "Hispanic/Latina/e/o", White_HispLatinx = C("White;Hispanic/Latina/e/o", "Hispanic/Latina/e/o;White"), Black_HispLatinx = "Black, African American, or African;Hispanic/Latina/e/o", .default = "2+ Races"))

```

```{r}
#one of the UDEs is DOB. Possibility that these dates are being inputted as the current date as a way to bypass the system

data %>% filter(Age.at.Admission == 0) %>% count(Program, sort = TRUE)
data %>% filter(Age.at.Admission == 0) %>% ggplot(aes(x = Length.of.Program.Stay)) + geom_histogram(binwidth = 5) + xlim(0, 365)

data %>% filter(Age.at.Admission == 120) %>% count(Program, sort = TRUE)
data %>% filter(Age.at.Admission == 121) %>% count(Program, sort = TRUE)
data %>% filter(Age.at.Admission == 122) %>% count(Program, sort = TRUE)

data %>% filter(Age.at.Admission <= 10 & Relationship.to.Primary.Client == "Self") %>% count(Program)
data %>% filter(Age.at.Admission <= 10 & Relationship.to.Primary.Client == "Self") %>% select()
householdvector <- c("408997", "H290580041", "402507", "402508", "402510", "H413945041", "402518", "402504", "199817", "H415618041", "H415623041", "405587", "397164", "402505", "401675", "404000", "H399886041", "H399891041", "H399892041", "401461", "393377", "402545")
data %>% filter(Global.Household.ID %in% householdvector) %>% count(Global.Household.ID)

data %>% filter(Gender..HMIS. == "Data not collected" | Gender..HMIS. == "") %>% count(Program, sort = TRUE)
data %>% filter(Gender..HMIS. == "Data not collected" | Gender..HMIS. == "") %>% ggplot(aes(x = Length.of.Program.Stay)) + geom_histogram(binwidth = 5) + xlim(0, 365)

data %>% filter(Enrollment.CoC..Admission. == "") %>% count(Program, sort = TRUE)
data %>% filter(Enrollment.CoC..Admission. == "NY-000 - No Assigned CoC") %>% count(Program, sort = TRUE)

data %>% count(Relationship.to.Primary.Client, sort = TRUE)
```



